---
title: "Reproducible Research - Impact of weather events on public health and economy across USA"
author: "Vince Force"
date: "27 décembre 2017"
output:
  html_document:
    toc: yes
    toc_depth: 5
    template: MyRenderingTemplate.html
  md_document:
    toc: yes
    toc_depth: 5
  pdf_document:
    toc: yes
    toc_depth: 5
mainfont: Arial
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

# Synopsis

The meaning of this analysis is to give event types with greater impacts on health and economy.

After a global point of view considering all data across USA, we give event types with greater impacts by state and by county.

The analysis uses a cluster classification of event types.

It also takes care of selecting accurate dates, as older dates contain less data.

# ResourcesSum of impacts

## R libraries

The libraries loaded for analysis are R.utils, plyr, dlpyr, reshape2, lattice, knitr and kableExtra.

```{r loadlibraries, message=FALSE}
library(R.utils)
library(plyr)
library(dplyr)
library(reshape2)
library(lattice)
library(knitr)
library(kableExtra)
```

## Custom functions

The custom functions extractYear, topClustersByState and topClustersByCounty will be used.

```{r customfunctions}
extractYear <- function(stringColValue) {
    splt <- strsplit(stringColValue, " ")
    format(as.Date(splt[[1]][1], format="%m/%d/%Y"),"%Y")
}
# top 4
topClustersByState <- function(stateCode, ds) {
    
    dsState  <- ds[ds$STATE == stateCode,]
    dsStateMelt <- melt(dsState, id = c("cluster"), measure.vars = c("impacts"))
    
    dsStateMeltSum <- dcast(dsStateMelt, cluster ~ variable, sum)
    dsStateMeltSum$cluster <- as.numeric(dsStateMeltSum$cluster)
    
    dsStateMeltMean <- dcast(dsStateMelt, cluster ~ variable, mean)
    dsStateMeltMean$cluster <- as.numeric(dsStateMeltMean$cluster)
    
    dsStateMeltData <- dsStateMeltSum
    dsStateMeltData$impacts <- round(dsStateMeltData$impacts, 2)
    dsStateMeltData$meanimpacts <- round(dsStateMeltMean$impacts, 2)

    odsStateMeltData <- arrange(dsStateMeltData, -impacts)
    odsStateMeltDataTop <- odsStateMeltData[1:4,]
    
    tCBS <- data.frame(
        cl1 = odsStateMeltDataTop$cluster[1],
        sm1 = odsStateMeltDataTop$impacts[1],
        mn1 = odsStateMeltDataTop$meanimpacts[1],
        cl2 = odsStateMeltDataTop$cluster[2],
        sm2 = odsStateMeltDataTop$impacts[2],
        mn2 = odsStateMeltDataTop$meanimpacts[2],
        cl3 = odsStateMeltDataTop$cluster[3],
        sm3 = odsStateMeltDataTop$impacts[3],
        mn3 = odsStateMeltDataTop$meanimpacts[3],
        cl4 = odsStateMeltDataTop$cluster[4],
        sm4 = odsStateMeltDataTop$impacts[4],
        mn4 = odsStateMeltDataTop$meanimpacts[4]
    )

    tCBS
}

topClustersByCounty <- function(CountyName, state, ds) {
    
    dsCounty <- ds[ds$COUNTYNAME == CountyName & ds$STATE == state,]
    dsCountyMelt <- melt(dsCounty, id = c("cluster"), measure.vars = c("impacts"))

    dsCountyMeltSum <- dcast(dsCountyMelt, cluster ~ variable, sum)
    
    dsCountyMeltMean <- dcast(dsCountyMelt, cluster ~ variable, mean)

    dsCountyMeltData <- dsCountyMeltSum
    dsCountyMeltData$impacts <- round(dsCountyMeltData$impacts, 2)
    dsCountyMeltData$meanimpacts <- round(dsCountyMeltMean$impacts, 2)

    odsCountyMeltData <- arrange(dsCountyMeltData, -impacts)
    odsCountyMeltDataTop <- odsCountyMeltData[1:4,]
    
    tCBC <- data.frame(
        cl1 = odsCountyMeltDataTop$cluster[1],
        sm1 = odsCountyMeltDataTop$impacts[1],
        mn1 = odsCountyMeltDataTop$meanimpacts[1],
        cl2 = odsCountyMeltDataTop$cluster[2],
        sm2 = odsCountyMeltDataTop$impacts[2],
        mn2 = odsCountyMeltDataTop$meanimpacts[2],
        cl3 = odsCountyMeltDataTop$cluster[3],
        sm3 = odsCountyMeltDataTop$impacts[3],
        mn3 = odsCountyMeltDataTop$meanimpacts[3],
        cl4 = odsCountyMeltDataTop$cluster[4],
        sm4 = odsCountyMeltDataTop$impacts[4],
        mn4 = odsCountyMeltDataTop$meanimpacts[4]
    )

    tCBC
}
```

## Rmd document - R code
The original Rmd document with all R code can be found at [https://github.com/vinceforce/datasciencecoursera/blob/master/CourseProject2.Rmd].

# Data processing

## Data loading

Data is downloaded from the following URL [https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2].

The dataset is unzipped to a CSV file.

NB : this part needs package R.utils to unzip the bz2 file. Use install.packages("R.utils") if it is not installed.

```{r loaddata, cache=TRUE}
# lsDir <- dir()
if (!"data" %in% dir()) {
        dir.create("data")
        dataURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
        download.file(dataURL, "data/repdata%2Fdata%2FStormData.csv.bz2", method = "libcurl", mode = "wb")
        bunzip2(paste(getwd(), "data/repdata%2Fdata%2FStormData.csv.bz2", sep = "/"), paste(getwd(), "data/repdata%2Fdata%2FStormData.csv", sep = "/"), remove = FALSE, skip = TRUE)
        unlink("data/repdata%2Fdata%2FStormData.csv.bz2")
}
# lsGlobEnv <- ls()
if (!"StormData" %in% ls()) {
        ## This first line will likely take a few seconds. Be patient!
        StormData <- read.csv("data/repdata%2Fdata%2FStormData.csv", stringsAsFactors = FALSE)
}
```

## Data preparation

A quick look to dataset content (str, head) for identifying columns to be considered for analysis.

```{r preparedata-1, cache=TRUE}
str(StormData)
head(StormData)
```

2 new datasets are created from original StormData dataset, one for each question:

- sdHealth with only useful and mostly valued columns (EVTYPE, FATALITIES, INJURIES, BGN_DATE, STATE, COUNTYNAME)
- sdEconom with only useful and mostly valued columns (EVTYPE, PROPDMG, CROPDMG, BGN_DATE, STATE, COUNTYNAME)

FATALITIES and INJURIES are integers, PROPDMG and CROPDMG are numbers (in k$).
EVTYPE, BGN_DATE, STATE, COUNTYNAME are strings (BGN_DATE will be processed to extract year).

```{r preparedata-2, cache=TRUE}
sdHealth <- StormData[
    complete.cases("EVTYPE", "BGN_DATE", "COUNTYNAME", "STATE")
        & (StormData$INJURIES > 0 | StormData$FATALITIES > 0),
    c("EVTYPE", "FATALITIES", "INJURIES", "BGN_DATE", "COUNTYNAME", "STATE")]
sdHealth$FATALITIES[is.na(sdHealth$FATALITIES) | is.nan(sdHealth$FATALITIES)] <- 0
sdHealth$FATALITIES[is.na(sdHealth$INJURIES) | is.nan(sdHealth$INJURIES)] <- 0
sdEconom <- StormData[
    complete.cases("EVTYPE", "BGN_DATE", "COUNTYNAME", "STATE")
        & (StormData$PROPDMG > 0 | StormData$CROPDMG > 0),
    c("EVTYPE", "PROPDMG", "CROPDMG", "BGN_DATE", "COUNTYNAME", "STATE")]
sdEconom$PROPDMG[is.na(sdEconom$PROPDMG) | is.nan(sdEconom$PROPDMG)] <- 0
sdEconom$CROPDMG[is.na(sdEconom$CROPDMG) | is.nan(sdEconom$CROPDMG)] <- 0
```

## Data preparation for health impact analysis
A cumulative impacts column is added, sum of FATALITIES and INJURIES.

For calculation optimization, we discard records with 0 value for impacts.

From events date, calculation of event year, using extractYear custom function.

First ranking of event types, with all data.

```{r healthanalysis, cache=TRUE}
sdHealth$impacts <- sdHealth$FATALITIES + sdHealth$INJURIES
sdHealth$begYear <- sapply(sdHealth$BGN_DATE, extractYear)

ssH <- tapply(sdHealth$impacts, sdHealth$EVTYPE, sum, na.rm = TRUE)
ssHMean <- tapply(sdHealth$impacts, sdHealth$EVTYPE, mean, na.rm = TRUE)
dfH <- data.frame(EVTYPE = rownames(ssH), impacts = ssH, meanimpacts = ssHMean)
odfH <- arrange(dfH, -impacts)
kable(head(odfH, 50)) %>% kable_styling(bootstrap_options = c("striped"))
maxSSH <- max(ssH)
```

## Data preparation for economy impact analysis
A cumulative impacts column is added, sum of PROPDMG and CROPDMG.

For calculation optimization, we discard records with 0 value for impacts.

From events date, calculation of event year, using extractYear custom function.

First ranking of event types, with all data.

```{r economyanalysis, cache=TRUE}
sdEconom$impacts <- sdEconom$PROPDMG + sdEconom$CROPDMG
sdEconom$begYear <- sapply(sdEconom$BGN_DATE, extractYear)

ssE <- tapply(sdEconom$impacts, sdEconom$EVTYPE, sum, na.rm = TRUE)
ssEMean <- tapply(sdEconom$impacts, sdEconom$EVTYPE, mean, na.rm = TRUE)
dfE <- data.frame(EVTYPE = rownames(ssE), impacts = ssE, meanimpacts = ssEMean)
odfE <- arrange(dfE, -impacts)
kable(head(odfE, 50)) %>% kable_styling(bootstrap_options = c("striped"))
maxSSE <- max(ssE)

```

## Event types clustering
Event types labels appear to be entered manually, with varying values for same event type (eg : TORNADO, TRONADO, TORNADOES, etc.).

### Clusters calculation
We use hierarchic clustering to group similar event type labels.

We cut at 2/3 of the number of distincts EVTYPEs.

```{r eventtypeclustering, fig.height=8, out.width="80%"}
EvtTypesH <- unique(as.character(sdHealth$EVTYPE))
d <- adist(EvtTypesH, ignore.case = TRUE)
rownames(d) <- EvtTypesH
dd <- as.dist(d)
hcH <- hclust(as.dist(d))
klimH = floor(2 * length(EvtTypesH) / 3)
clustersH <- data.frame(cutree(hcH, k = klimH))
clustersH$EVTYPE <- rownames(clustersH)
names(clustersH) <- c("cluster", "EVTYPE")
oclustersH <- arrange(clustersH, cluster)

EvtTypesE <- unique(as.character(sdEconom$EVTYPE))
d <- adist(EvtTypesE, ignore.case = TRUE)
rownames(d) <- EvtTypesE
dd <- as.dist(d)
hcE <- hclust(as.dist(d))
klimE = floor(2 * length(EvtTypesE) / 3)
clustersE <- data.frame(cutree(hcE, k = klimE))
clustersE$EVTYPE <- rownames(clustersE)
names(clustersE) <- c("cluster", "EVTYPE")
oclustersE <- arrange(clustersE, cluster)

par(mfrow = c(2, 1))
plot(hcH, cex = 0.3, main = "Event types hierarchical classification dendogram - Health",
             xlab = "EVTYPE")
rect.hclust(hcH, k = klimH)
plot(hcE, cex = 0.3, main = "Event types hierarchical classification dendogram - Economy",
             xlab = "EVTYPE")
rect.hclust(hcE, k = klimE)
par(mfrow = c(1, 1))
```

For having an idea about clustering results, let us print the 100 first values of event types, ordered by cluster ID.

#### Event types cluster classification - Health
```{r eventtypesHList}
kable(head(oclustersH, 100)) %>% kable_styling(bootstrap_options = c("striped"))
```

#### Event types cluster classification - Economy
```{r eventtypesEList}
kable(head(oclustersE, 100)) %>% kable_styling(bootstrap_options = c("striped"))
```

### Clusters attachment
We join cluster column to health and economy datasets.

```{r clusterjoin, message=FALSE}
sdHealthC <- left_join(sdHealth, clustersH)
sdEconomC <- left_join(sdEconom, clustersE)

evtTypesCluster <- function(clusters) {
    cls <- unique(clusters$cluster)
    typesList <- data.frame()
    for (cl in cls) {
        typesArr <- clusters[clusters$cluster == cl, "EVTYPE"]
        listT <- ""
        beg = TRUE
        for (type in typesArr) {
            if (beg) {
                listT <- type
                beg = FALSE
            }
            else {
                listT <- paste(listT, type, sep = ", ")
            }
            
        }
        typesList <- rbind(typesList, data.frame(cluster = cl, types = listT))
    }
    typesList
}

typeslistH <- evtTypesCluster(clustersH)
clustersTH <- left_join(data.frame(cluster = unique(clustersH$cluster)), typeslistH)
typeslistE <- evtTypesCluster(clustersE)
clustersTE <- left_join(data.frame(cluster = unique(clustersE$cluster)), typeslistE)
```

# Event types by cluster - Health analysis

The following table gives the event types linked to each cluster.

```{r clustertypeshealth}
kable(clustersTH) %>% kable_styling(bootstrap_options = c("striped"))
```

# Event types by cluster - Economy analysis

The following table gives the event types linked to each cluster.

```{r clustertypeseconomy}
kable(clustersTE) %>% kable_styling(bootstrap_options = c("striped"))
```

# Impacts by cluster
```{r healthclusteranalysis}
ssHC <- tapply(sdHealthC$impacts, as.factor(sdHealthC$cluster), sum, na.rm = TRUE)
dfHC <- data.frame(cluster = as.numeric(rownames(ssHC)), impacts = ssHC)
ssHCMean <- tapply(sdHealthC$impacts, as.factor(sdHealthC$cluster), mean, na.rm = TRUE)
dfHCMean <- data.frame(cluster = as.numeric(rownames(ssHCMean)), impacts = ssHCMean)
```
```{r economyclusteranalysis}
ssEC <- tapply(sdEconomC$impacts, as.factor(sdEconomC$cluster), sum, na.rm = TRUE)
dfEC <- data.frame(cluster = as.numeric(rownames(ssEC)), impacts = ssEC)
ssECMean <- tapply(sdEconomC$impacts, as.factor(sdEconomC$cluster), mean, na.rm = TRUE)
dfECMean <- data.frame(cluster = as.numeric(rownames(ssECMean)), impacts = ssECMean)
sdEconomCDD <- sdEconomC[sdEconomC$begYear >= 1992,]
ssECDD <- tapply(sdEconomCDD$impacts, as.factor(sdEconomCDD$cluster), sum, na.rm = TRUE)
```

For each question, we draw a barplot showing impacts sum by cluster ID, and a barplot showing impacts mean by cluster ID.

```{r barplots, fig.height=8, out.width="80%"}
par(mfrow = c(2, 2))
 
barplot(t(as.matrix(dfHC)), horiz = FALSE, col = "red",
        main = "Health impacts - sum",
        xlab = "Cluster ID", ylab = "Sum of impacts (fatalities + injuries)")

barplot(t(as.matrix(ssHCMean)), horiz = FALSE, col = "red",
        main = "Health impacts - mean",
        xlab = "Cluster ID", ylab = "Mean of impacts (fatalities + injuries)")

barplot(t(as.matrix(dfEC)), horiz = FALSE, col = "red",
        main = "Economy impacts - sum",
        xlab = "Cluster ID", ylab = "Sum of impacts (prop + crop), in k$")

barplot(t(as.matrix(dfECMean)), horiz = FALSE, col = "red",
        main = "Economy impacts - mean",
        xlab = "Cluster ID", ylab = "Mean of impacts (prop + crop), in k$")
```

If the ranking by sum differ from the ranking by mean (not considering NA values), it can mean that we have to reduce the analysis date range, but not too much as very rare event types with big mean are not relevant.

For health analysis, the ranking stays almost the same.

For econmy analysis, the ranking differs.
 
For proceeding with global analysis, we keep the ranking of clusters with all date by sum to select the most impactant clusters, but we will then restrict the date range and see how it impacts the ranking by sum.

For proceeding with analysis by state and county, we will always consider the full range for dates.

## With all dates considered
In order to find the clusters with the greatest impacts, we order the dataframe by impacts column value (decreasing).

Then we list the event types associated with each cluster for the 6 first clusters.

### Cluster ranking for health impacts

We list the 6 first clusters with greater impacts, with cluster DI, sum of impacts, and the list of event types associated with this cluster.

```{r}
odfHDD <- arrange(dfHC, -impacts)
odfHDD$cluster <- as.numeric(odfHDD$cluster)
# head(odfHDD)
maxSSHDD <- max(ssHC)
indexClustMaxHDD <- which(ssHC == maxSSHDD)
clustMaxHDD <- ssHC[indexClustMaxHDD]
clustNumberHDD <- names(ssHC)[indexClustMaxHDD]
EVTYPESMaxHDD <- clustersH[which(clustersH$cluster == as.numeric(clustNumberHDD)), c("EVTYPE")]
# EVTYPESMaxHDD

legendH <- vector(length = 6)
for (indClust in 1:6) {
    clusterID <- as.numeric(odfHDD$cluster[indClust])
    print(paste("clusterID", clusterID, sep = " : "))
    clusterImpacts <- odfHDD$impacts[indClust]
    print(paste("Sum of impacts", clusterImpacts, sep = " : "))
    EVTYPES <- clustersH[which(clustersH$cluster == as.numeric(clusterID)), c("EVTYPE")]
    print(paste("EVTYPE", EVTYPES, sep = " : "))
    print("------------------------------------------------------")
    evttypesList <- ""
    evtypeslisted <- 0
    for (et in EVTYPES) {
        evttypesList <- paste(evttypesList, et, sep = "")
        evtypeslisted <- evtypeslisted + 1
        if (evtypeslisted < length(EVTYPES)) {
            evttypesList <- paste(evttypesList, ", ", sep = "")
        }
    }
    legendH[indClust] <- paste(as.character(clusterID), evttypesList, sep = " : ")
}

sdHealthHead <- sdHealthC[as.numeric(sdHealthC$cluster) %in% as.numeric(odfHDD$cluster[1:6]),]
sdHealthHead$cluster <- as.factor(sdHealthHead$cluster)
sdHealthHead$impacts <- as.numeric(sdHealthHead$impacts)
sdHealthHead$begYear <- as.numeric(sdHealthHead$begYear)

sdHealthMelt <- melt(sdHealthHead, id = c("cluster", "begYear"), measure.vars = c("impacts"))
sdHealthMeltData <- dcast(sdHealthMelt, begYear + cluster ~ variable, sum)

```

### Cluster ranking for economy impacts

We list the 6 first clusters with greater impacts, with cluster DI, sum of impacts (in k$), and the list of event types associated with this cluster.

```{r}
odfEDD <- arrange(dfEC, -impacts)
odfEDD$cluster <- as.numeric(odfEDD$cluster)
# head(odfEDD)
maxSSEDD <- max(ssEC)
indexClustMaxEDD <- which(ssEC == maxSSEDD)
clustMaxEDD <- ssEC[indexClustMaxEDD]
clustNumberEDD <- names(ssEC)[indexClustMaxEDD]
EVTYPESMaxEDD <- clustersE[which(clustersE$cluster == as.numeric(clustNumberEDD)), c("EVTYPE")]
# EVTYPESMaxEDD
```

#### With discarded data

We list the 6 first clusters with greater impacts, with cluster DI, sum of impacts (in k$), and the list of event types associated with this cluster.

```{r}
odfE <- arrange(dfEC[dfEC$begYear >= 1992,], -impacts)
odfE$cluster <- as.numeric(odfE$cluster)
# head(odfE)
maxSSE <- max(ssECDD)
indexClustMaxE <- which(ssECDD == maxSSE)
clustMaxE <- ssEC[indexClustMaxE]
clustNumberE <- names(ssECDD)[indexClustMaxE]
EVTYPESMaxE <- clustersE[which(clustersE$cluster == as.numeric(clustNumberE)), c("EVTYPE")]
# EVTYPESMaxE

legendE = vector(length = 6)
for (indClust in 1:6) {
    clusterID <- as.numeric(odfEDD$cluster[indClust])
    print(paste("clusterID", clusterID, sep = " : "))
    clusterImpacts <- odfEDD$impacts[indClust]
    print(paste("Sum of impacts", clusterImpacts, sep = " : "))
    EVTYPES <- clustersE[which(clustersE$cluster == as.numeric(clusterID)), c("EVTYPE")]
    print(paste("EVTYPE", EVTYPES, sep = " : "))
    print("------------------------------------------------------")
    evttypesList <- ""
    evtypeslisted <- 0
    for (et in EVTYPES) {
        evttypesList <- paste(evttypesList, et, sep = "")
        evtypeslisted <- evtypeslisted + 1
        if (evtypeslisted < length(EVTYPES)) {
            evttypesList <- paste(evttypesList, ", ", sep = "")
        }
    }
    leg <- paste(as.character(clusterID), evttypesList, sep = " : ")
    legS <- ""
    beginning <- TRUE
    limCharLegend <- 80
    if (nchar(leg)[1] > limCharLegend) {
        legparts <- strsplit(leg, ", ")
        if (indClust == 4) {
        }
        nbChar <- 0
        for (part in legparts[[1]]) {
            if (beginning) {
                legS <- part
                nbChar <- nchar(part)
                beginning <- FALSE
            }
            else {
                nbChar <- nbChar + nchar(part)
                if(nbChar > limCharLegend) {
                        legS <- paste(legS, part, sep = "\n\t")
                        nbChar <- nchar(part)
                }
                else {
                    legS <- paste(legS, part, sep = ", ")
                }
            }
        }
    }
    else {
        legS <- leg
    }
    legendE[indClust] <- legS
}

sdEconomHead <- sdEconomC[as.numeric(sdEconomC$cluster) %in% as.numeric(odfEDD$cluster[1:6]),]
sdEconomHead$cluster <- as.factor(sdEconomHead$cluster)
sdEconomHead$impacts <- as.numeric(sdEconomHead$impacts)
sdEconomHead$begYear <- as.numeric(sdEconomHead$begYear)

sdEconomMelt <- melt(sdEconomHead, id = c("cluster", "begYear"), measure.vars = c("impacts"))
sdEconomMeltData <- dcast(sdEconomMelt, begYear + cluster ~ variable, sum)
```


## Comparison with discarding dates with no data
We draw a multipanel plot to compare the 6 first clusters, presenting, for each cluster, sums by year (plots), mean of these values accross years (horizontal line and caption) and sum of these values accross years (caption).
The plot part for all dates indicates the dates to be considered as a floor:
- health -> 1982
- econmy -> 1992
These date floors are used in the other plots (legended with meaningful values).

```{r clusteranalysiscontinued, fig.height=16, cache=TRUE, out.width="80%", fig.align='center'}
gH <- xyplot(impacts ~ begYear | cluster, data = sdHealthMeltData,
       main = list(label = "Impacts on public health", cex = 0.9),
       xlab = "Begin year",
       ylab = "Injuries + fatalities",
       panel = function(x, y) {
                panel.xyplot(x, y, col = "green", pch = 19)
                panel.abline(h = mean(y), col = "red")
                captionSum <- paste("Sum", round(sum(y), 2), sep = " : ")
                captionMean <- paste("Mean", round(mean(y), 2), sep = " : ")
                caption <- paste(captionSum, captionMean, sep = " - ")
                panel.text(caption, x = 1980, y = 3 * mean(y), cex = 0.8)
           },
       key=list(
           title = "",
           space = "bottom",
           text=list(lab=legendH, cex=.6)
           )
       )
gHDD <- xyplot(impacts ~ begYear | cluster,
               data = sdHealthMeltData[sdHealthMeltData$begYear >= 1982,],
       main = list(label = "Impacts on public health - Meaningful dates", cex = 0.9),
       xlab = "Begin year",
       ylab = "Injuries + fatalities",
       panel = function(x, y) {
                panel.xyplot(x, y, col = "green", pch = 19)
                panel.abline(h = mean(y), col = "red")
                captionSum <- paste("Sum", round(sum(y), 2), sep = " : ")
                captionMean <- paste("Mean", round(mean(y), 2), sep = " : ")
                caption <- paste(captionSum, captionMean, sep = " - ")
                panel.text(caption, x = 1995, y = 3 * mean(y), cex = 0.8)
           },
       key=list(
           title = "",
           space = "bottom",
           text=list(lab=legendH, cex=.6)
           )
       )

gE <- xyplot(1e-3 * impacts ~ begYear | cluster, data = sdEconomMeltData,
       main = list(label = "Impacts on economy", cex = 0.9),
       xlab = "Begin year",
       ylab = "Property + crop damages (in M$)",
       panel = function(x, y) {
                panel.xyplot(x, y, col = "green", pch = 19)
                panel.abline(h = mean(y), col = "red")
                captionSum <- paste("Sum", round(sum(y), 0), sep = " : ")
                captionMean <- paste("Mean", round(mean(y), 0), sep = " : ")
                caption <- paste(captionSum, captionMean, sep = " - ")
                panel.text(caption, x = 1980, y = 1.1 * mean(y), cex = 0.8)
           },
       key=list(
           title = "",
           space = "bottom",
           text=list(lab=legendE, cex=.6)
           )

       )

gEDD <- xyplot(1e-3 * impacts ~ begYear | cluster,
               data = sdEconomMeltData[sdEconomMeltData$begYear >= 1992,],
       main = list(label = "Impacts on economy - Meaningful dates", cex = 0.9),
       xlab = "Begin year",
       ylab = "Property + crop damages (in M$)",
       panel = function(x, y) {
                panel.xyplot(x, y, col = "green", pch = 19)
                panel.abline(h = mean(y), col = "red")
                captionSum <- paste("Sum", round(sum(y), 0), sep = " : ")
                captionMean <- paste("Mean", round(mean(y), 0), sep = " : ")
                caption <- paste(captionSum, captionMean, sep = " - ")
                panel.text(caption, x = 2000, y = 1.1 * mean(y), cex = 0.8)
           },
       key=list(
           title = "",
           space = "bottom",
           text=list(lab=legendE, cex=.6)
           )

       )
plot(gH, split = c(1, 1, 1, 4), more = TRUE)
plot(gHDD, split = c(1, 2, 1, 4), more = TRUE)
plot(gE, split = c(1, 3, 1, 4), more = TRUE)
plot(gEDD, split = c(1, 4, 1, 4))
```

# Global impacts

## Global impact on public heath

When we consider all dates, the cluster with the maximum impact on public health is number **`r clustNumberHDD`**.

The list of EVTYPE associated with this cluster is `r EVTYPESMaxHDD`.

As we look at the mean values, the ranking stays the same.

As we dicard dates with partial data (prior to 1982), the results stay the same.

## Global impact on economy

When we consider all dates, the cluster with the greatest economic impact is number **`r clustNumberEDD`**.

The list of EVTYPE associated with this cluster is `r EVTYPESMaxEDD`.

As we look at the mean values, the ranking changes. Some other event types clusters have a greater mean value.

Logically, as we dicard dates with partial data (prior to 1992), the results change, and even the ranking on sums changes. Some other event types clusters have a greater sum value.

The cluster with the greatest economic impact is number **`r clustNumberE`**.

The list of EVTYPE associated with this cluster is`r EVTYPESMaxE`.

# Results

## Impacts  by state

### List of states

For a better understanding, a list of state names is bounded to the data set (available on [https://github.com/vinceforce/datasciencecoursera/blob/master/USAstates.csv.bz2]).

```{r lodadatastates, message=FALSE, cache=TRUE}
if (!"data" %in% dir()) {
        dir.create("data")
        dataURL <- "https://github.com/vinceforce/datasciencecoursera/blob/master/USAstates.csv.bz2"
        download.file(dataURL, "data/USAstates.csv.bz2", method = "libcurl", mode = "wb")
        bunzip2(paste(getwd(), "data/USAstates.csv.bz2", sep = "/"), paste(getwd(), "data/USAstates.csv", sep = "/"), remove = FALSE, skip = TRUE)
        unlink("data/USAstates.csv.bz2")
}
# lsGlobEnv <- ls()
if (!"USAstateslist" %in% ls()) {
        USAstateslist <- read.csv("data/USAstates.csv", stringsAsFactors = FALSE)
}

sdHealthHeadS <- left_join(sdHealthHead, USAstateslist)
sdEconomHeadS <- left_join(sdEconomHead, USAstateslist)
sdHealthCS <- left_join(sdHealthC, USAstateslist)
sdEconomCS <- left_join(sdEconomC, USAstateslist)
```

### Impact on public heath

#### 4 factors
The folowing table gives, for each state, the 4 greater impacts with cluster id, sum and mean (of the sums by year discarding NA values for year) by cluster for this state.

To get the list of event types for one cluster ID, please refer to chapter above giving this list.

```{r, message=FALSE, warning=FALSE, cache=TRUE}
sdHealthCS <- sdHealthCS[as.numeric(sdHealthC$begYear) >= 1980,]
sdHealthCSM <- sdHealthCS
stateCodes <- unique(sdHealthCSM$STATE)
states <- left_join(data.frame(STATE = stateCodes), USAstateslist)

dfHByState <- data.frame()
for (indstate in 1:nrow(states)) {
    state <- states[indstate, "STATE"]
    stName <- states[indstate, "stateName"]
    row = data.frame(STATE = state, stateName = stName)
    dfHByState <- rbind(dfHByState, cbind(row, topClustersByState(state, sdHealthCSM)))
}
kable(arrange(dfHByState, STATE)) %>% kable_styling(bootstrap_options = c("striped"))
```

#### 2 factors with EVTYPE labels
The folowing table gives, for each state, the 2 greater impacts with the list of event types linked with cluster, sum and mean (of the sums by year discarding NA values for year) by cluster for this state.
```{r, message=FALSE, warning=FALSE, cache=TRUE}
dfHByStateRed <- dfHByState[ ,1:8]
dfHBSR1 <- left_join(dfHByStateRed, typeslistH, by = c("cl1" = "cluster"))
colnames(dfHBSR1)[9] = "Types1"
dfHBSRT <- left_join(dfHBSR1, typeslistH, by = c("cl2" = "cluster"))
colnames(dfHBSRT)[10] = "Types2"
dfHBSRT <- dfHBSRT[ ,c(1:2, 9, 4:5, 10, 7:8)]
kable(arrange(dfHBSRT, STATE)) %>% kable_styling(bootstrap_options = c("striped")) %>% column_spec(3, width = "10em")
```


### Impact on economy

#### 4 factors
The folowing table gives, for each state, the 4 greater impacts with cluster id, sum and mean (of the sums by year discarding NA values for year) by cluster for this state.

To get the list of event types for one cluster ID, please refer to chapter above giving this list.

```{r, message=FALSE, warning=FALSE, cache=TRUE}
sdEconomCS <- sdEconomCS[as.numeric(sdEconomCS$begYear) >= 1992,]
sdEconomCSM <- sdEconomCS
sdEconomCSM$impacts <- 1e-3 * sdEconomCSM$impacts
stateCodes <- unique(sdEconomCS$STATE)
states <- left_join(data.frame(STATE = stateCodes), USAstateslist)
dfEByState <- data.frame()
for (indstate in 1:nrow(states)) {
    state <- states[indstate, "STATE"]
    stName <- states[indstate, "stateName"]
    row = data.frame(STATE = state, stateName = stName)
    dfEByState <- rbind(dfEByState, cbind(row, topClustersByState(state, sdEconomCSM)))
}
kable(arrange(dfEByState, STATE)) %>% kable_styling(bootstrap_options = c("striped"))
```

#### 2 factors with EVTYPE labels
The folowing table gives, for each state, the 2 greater impacts with the list of event types linked with cluster, sum and mean (of the sums by year discarding NA values for year) by cluster for this state.
```{r, message=FALSE, warning=FALSE, cache=TRUE}
dfEByStateRed <- dfEByState[ ,1:8]
dfEBSR1 <- left_join(dfEByStateRed, typeslistE, by = c("cl1" = "cluster"))
colnames(dfEBSR1)[9] = "Types1"
dfEBSRT <- left_join(dfEBSR1, typeslistE, by = c("cl2" = "cluster"))
colnames(dfEBSRT)[10] = "Types2"
dfEBSRT <- dfEBSRT[ ,c(1:2, 9, 4:5, 10, 7:8)]
kable(arrange(dfEBSRT, STATE)) %>% kable_styling(bootstrap_options = c("striped")) %>% column_spec(3, width = "10em")
```

## Impacts  by county

### Impact on public heath

#### Data preparation
As we look at COUNTYNAME column, we see there is a a lot of strange values like "AZO16" or "AZO16 > AZ028". We decide to discard the corresponding records with a regular expression eliminating values with digits in.

Calculation times will also be less important this way.

#### 4 factors
The folowing table gives, for each county, the 4 greater impacts with cluster id, sum and mean (of the sums by year discarding NA values for year) by cluster for this county.

To get the list of event types for one cluster ID, please refer to chapter above giving this list.

```{r countyhealth, warning=FALSE, cache=TRUE}
sdHealthCSM <- sdHealthCS[
    sdHealthCS$COUNTYNAME %in% grep("^([^0-9])*$", sdHealthCS$COUNTYNAME,
                                    value = TRUE, ignore.case = TRUE),]
sdHealthCSM$COUNTYNAME <- strtrim(sdHealthCSM$COUNTYNAME, 30)
counties <- unique(sdHealthCSM[, c("COUNTYNAME", "STATE")])

dfHByCounty <- data.frame()
for (indCounty in 1:nrow(counties)) {
    county <- counties[indCounty, "COUNTYNAME"]
    state <- counties[indCounty, "STATE"]
    row = data.frame(COUNTYNAME = county, STATE = state)
    dfHByCounty <- rbind(dfHByCounty, cbind(row, topClustersByCounty(county, state, sdHealthCSM)))
}
dfHByCounty$COUNTYNAME <- as.character(dfHByCounty$COUNTYNAME)
kable(arrange(dfHByCounty, COUNTYNAME)) %>% kable_styling(bootstrap_options = c("striped"))
```

#### 2 factors with EVTYPE labels
The folowing table gives, for each county, the 2 greater impacts with the list of event types linked with cluster, sum and mean (of the sums by year discarding NA values for year) by cluster for this county.
```{r, message=FALSE, warning=FALSE, cache=TRUE}
dfHByCountyRed <- dfHByCounty[ ,1:8]
dfHBCR1 <- left_join(dfHByCountyRed, typeslistH, by = c("cl1" = "cluster"))
colnames(dfHBCR1)[8] = "Types1"
dfHBCRT <- left_join(dfHBCR1, typeslistH, by = c("cl2" = "cluster"))
colnames(dfHBCRT)[9] = "Types2"
dfHBCRT <- dfHBCRT[ , c(1:2, 9, 4:5, 10, 7:8)]
kable(arrange(dfHBCRT, COUNTYNAME)) %>% kable_styling(bootstrap_options = c("striped")) %>% column_spec(3, width = "10em")
```

### Impact on economy

#### 4 factors
The folowing table gives, for each county, the 4 greater impacts with cluster id, sum and mean (of the sums by year discarding NA values for year) by cluster for this county.

To get the list of event types for one cluster ID, please refer to chapter above giving this list.

```{r countyeconmoy, warning=FALSE, cache=TRUE}
sdEconomCSM <- sdEconomCS[
    sdEconomCS$COUNTYNAME %in% grep("^([^0-9])*$", sdEconomCS$COUNTYNAME,
                                    value = TRUE, ignore.case = TRUE),]
sdEconomCSM$COUNTYNAME <- strtrim(sdEconomCSM$COUNTYNAME, 30)
counties <- unique(sdEconomCSM[,c("COUNTYNAME", "STATE")])

dfEByCounty <- data.frame()
for (countyInd in 1:nrow(counties)) {
    county <- counties[countyInd, "COUNTYNAME"]
    state <- counties[countyInd, "STATE"]
    row = data.frame(COUNTYNAME = county, STATE = state)
    dfEByCounty <- rbind(dfEByCounty, cbind(row, topClustersByCounty(county, state, sdEconomCSM)))
}
dfEByCounty$COUNTYNAME <- as.character(dfEByCounty$COUNTYNAME)
kable(arrange(dfEByCounty, COUNTYNAME)) %>% kable_styling(bootstrap_options = c("striped"))

```

#### 2 factors with EVTYPE labels
The folowing table gives, for each county, the 2 greater impacts with the list of event types linked with cluster, sum and mean (of the sums by year discarding NA values for year) by cluster for this county.
```{r, message=FALSE, warning=FALSE, cache=TRUE}
dfEByCountyRed <- dfEByCounty[ ,1:8]
dfEBCR1 <- left_join(dfEByCountyRed, typeslistE, by = c("cl1" = "cluster"))
colnames(dfEBCR1)[8] = "Types1"
dfEBCRT <- left_join(dfEBCR1, typeslistE, by = c("cl2" = "cluster"))
colnames(dfEBCRT)[9] = "Types2"
dfEBCRT <- dfEBCRT[ , c(1:2, 9, 4:5, 10, 7:8)]
kable(arrange(dfEBCRT, COUNTYNAME)) %>% kable_styling(bootstrap_options = c("striped")) %>% column_spec(3, width = "10em")
```


# Conclusion

The previous chapters gave global impacts of wheather event types on public health and econmy across the United States, and details by state and by county.

Analysis could be perfected by finding a more accurate way to group event types by cluster.

A more complete analysis could be performed for one state or one county taken apart.
