---
title: "Motor trend - MPG versus Transmission type"
author: "Vince Force"
date: "11 janvier 2018"
output:
  pdf_document: default
  html_document: default
geometry: "left=2cm,right=2cm,top=2cm,bottom=1.5cm"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, comment = "", warning = FALSE)
```

## Executive summary
We try to establish wether automatic transmission is better for mpg (miles per gallon), or not.

After exploratory analysis, we try to fit first a simple regression model, then a multivariate regression model, to establish the impact of transmission type, and estimate its value.

```{r dependencies, message=FALSE}
library(ggplot2)
library(car)
library(knitr)
library(ggplot2)
library(GGally)
```


## Exploratory analysis

The dataset is part of the "datasets" library. We have to load the library, then load the mtcars dataset.
```{r loaddata}
library(datasets)
data(mtcars)
```
We give in appendix some basic numbers about the dataset (dimensions, first 10 rows).
Using help(mtcars), we identify the column giving transmission type, it is am (0 = automatic, 1 = manual).
Finally we draw a boxplot giving global impact of transmission type on mpg distribution.

Considering the boxplot shown in Appendix, mpg seems to be higher as transmission is manual.
Let us further investigate.

## Data processing

We transform the appropriate columns into factors.

```{r dataprocessing}
c <- mtcars
c$am <- as.factor(c$am)
levels(c$am) <- c("auto", "manual")
c$cyl <- as.factor(c$cyl)
c$gear <- as.factor(c$gear) 
c$carb <- as.factor(c$carb)
c$vs <- as.factor(c$vs)
```

## Simple regression

### Analysis

We fit a simple regression model using am as predictor and mpg as outcome.
We perform a t-test for validating the hypothesis that mpg means differ following transmission type.
You will see in appendix that the residual error distrbution is almost normal, the residual errors are equally spaned around 0, and there is no heteroscedacity.
 
```{r simpleregression-analysis, echo = FALSE}
lmCar_am <- lm(mpg ~ am, data = c)
coeflmCar_am <- summary(lmCar_am)$coefficients
coeflmCar_am
tm <- as.character(lmCar_am$terms)
lmCar_amModel <- paste(tm[[2]], tm[[1]], tm[[3]])
t.test.am <- t.test(c[c$am == "manual", c("mpg")], c[c$am == "auto", c("mpg")], alternative = "two.sided", var.equal = FALSE, paired = FALSE)
conf.int.am <- t.test.am$conf.int
```


### Results
The p-value for the t-test is **`r round(t.test.am$p.value, 4)`** much less than 0.05 (cf appendix). The null hypothesis (equal means between automatic and manual transmission) is rejected.
**The transmission has an effect on mpg, manual transmission cars have globally a better mpg.**
According to this model, the estimated gain between automatic and manual transmission is **`r round(coeflmCar_am[2, 1], 2)` miles / gallon**. A 95% confidence interval for the impact is **`r round(conf.int.am, 2)`**.
The R-squared is **`r round(summary(lmCar_am)$r.squared, 2)`**.
The adjusted R-squared is **`r round(summary(lmCar_am)$adj.r.squared, 2)`**.
The residual standard error is **`r round(summary(lmCar_am)$sigma, 2)`**.


In order to adjust this gain value, we have now to consider a multivariate regression.

## Multivariate regression

### Select the accurate factors

We use the R step function to find an accurate multivariate model.

```{r lmstep, echo=TRUE}
lmStep <- step(lm(mpg ~ ., c), scope = list(lower = ~am), direction = "backward", trace = 0)
```



```{r nested-models-1}
tm <- as.character(lmStep$terms)
lmStepModel <- paste(tm[[2]], tm[[1]], tm[[3]])
```

The step function has selected the model **`r lmStepModel`**. For this model, here is the summary.

```{r dataanalysis-multivariate, fig.height=6}
summary(lmStep)
```

### Results
With the model **`r lmStepModel`** selected, the impact of am on mpg is **`r round(lmStep$coefficients["ammanual"], 2)`**, as other factors are fixed.
The R-squared is **`r round(summary(lmStep)$r.squared, 2)`**.
The adjusted R-squared is **`r round(summary(lmStep)$adj.r.squared, 2)`**.
The residual standard error is **`r round(summary(lmStep)$sigma, 2)`**.
In the residual plots given in appendix, we see there is no heteroscedacity (plots on the Scale-Location plot are spread along the x-axis) and the errors are spread equally around 0 (plot rResiduals vs Fitted). The distribution of errors is almost normal (Normal Q-Q plot). The plot Residuals vs Leverage shows there are no outliers (no point out of 0.5 Cook's distance zone).

# Conclusion

The positive impact of manual transmission versus automatic transmission has been evaluated follwing 2 models:

1. Simple regression **`r lmCar_amModel`** : impact is **`r round(coeflmCar_am[2, 1], 2)`** miles / gallon
2. Multivariate regression with formula **`r lmStepModel`** : impact is **`r round(lmStep$coefficients["ammanual"], 2)`** miles / gallon (with other factors fixed).


# Appendix

```{r appendix-dependencies, ref.label='dependencies', eval=FALSE}

```

```{r appendix-loaddata, ref.label='loaddata', eval=FALSE}

```



```{r appendix-exploratoryanalysis-2}
kable(head(mtcars), caption = "First rows of the mtcars dataset")
classes <- sapply(mtcars, class)
dfClasses <- data.frame()
dfClasses <- rbind(dfClasses, classes)
colnames(dfClasses) <- names(classes)
kable(dfClasses, caption = "Classes of dataset columns")
```

```{r exploratoryanalysis, echo=FALSE, fig.height=4}
m <- mtcars
m$amfactor <- as.factor(m$am)
levels(m$amfactor) = c("auto", "manual")
g <- ggplot(m, aes(x = amfactor, y = mpg))
g <- g + geom_boxplot(aes(fill = amfactor))
g <- g + xlab("Transmission")
g <- g + ylab("Miles/(US) gallon")
g <- g + ggtitle("MPG distribution vs Transmission type") 
g <- g + theme(plot.title = element_text(hjust = 0.5))
g
```

```{r appendix-dataprocessing, ref.label='dataprocessing', eval=FALSE}

```

```{r appendix-exploratoryanalysis, ref.label='exploratoryanalysis', eval=FALSE}

```

**_T-test for checking MPG means differ across transmission type_**

```{r appendix-simpleregression-analysis-ttest}
t.test.am
```
The p-value is much less than 0.05.

**_Residuals for simple regression_**

```{r appendix-simpleregression-analysis-resids, fig.height=3}
residCar_am <- residuals(lmCar_am)
fittedCar_am <- fitted.values(lmCar_am)
par(cex = 0.7, cex.axis = 1, cex.lab = 1, cex.main = 1)
par(mfrow = c(1, 2))
plot(fittedCar_am, residCar_am, xlab = "Predicted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = "dashed")
plot(density(residCar_am), xlab = "Residuals", ylab = "Density", main = "")
par(mfrow = c(1, 1))
```

**_Test for residuals normality - simple regression_**

```{r appendix-simpleregression-analysis, fig.height=3}
shapiro.test(lmCar_am$residuals)
```

```{r appendix-lmall-standarderrors, ref.label="lmall-standarderrors", eval=FALSE}

```


```{r appendix-nested-models-1, ref.label="nested-models-1", eval=FALSE}

```

**_Residuals for multivariate regression_**

```{r appendix-multiregression-resid, fig.height=3}
resid <- residuals(lmStep)
fitted <- fitted.values(lmStep)
par(mfrow = c(1,2))
plot(density(resid), xlab = "Residuals", ylab = "Density", main = "")
plot(fitted, resid, xlab = "Predicted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = "dashed")
shapiro.test(lmStep$residuals)
```

```{r appendix-multiregression-resid-2, fig.height=6}
par(mfrow = c(2,2))
plot(lmStep, cex = 0.8)
```

*Original Rmd document can be found here*

[https://github.com/vinceforce/datasciencecoursera/blob/master/mpg_vs_transmission.Rmd](https://github.com/vinceforce/datasciencecoursera/blob/master/mpg_vs_transmission.Rmd)
