---
title: "Motor trend - MPG versus Transmission"
author: "Vince Force"
date: "11 janvier 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```

## Executive summary
As a journalist for Motor Trend magazine, we try to establish wether automatic transmission is better for mpg (miles per gallon), or not.

After exploratory analysis, we try to fit first a simple regression model, then a multivariate regression model, to establish the impact of transmission type, and estimate its value.

```{r dependencies, message=FALSE}
library(ggplot2)
library(car)
library(kableExtra)
library(knitr)
library(ggplot2)
library(GGally)
library(lmtest)
```


## Exploratory analysis

The dataset is part of the "datasets" library. We have to load the library, then load the mtcars dataset.
```{r loaddata}
library(datasets)
data(mtcars)
```
We give in appendix some basic numbers about the dataset (dimensions, first 10 rows).
Using help(mtcars), we identify the column giving transmission type, it is am (0 = automatic, 1 = manual).
We check (see appendix) then the number of records per transmission type, for checking data is almost covering both cases.
Finally we draw a boxplot giving global impact of transmission type on mpg distribution.

Considering the boxplot shown in Appendix, mpg seems to be higher as transmission is manual.
Let us further investigate.

## Data processing

We transform the appropriate columns into factors.

```{r dataprocessing}
c <- mtcars
c$am <- as.factor(c$am)
levels(c$am) <- c("auto", "manual")
c$cyl <- as.factor(c$cyl)
c$gear <- as.factor(c$gear) 
c$carb <- as.factor(c$carb)
c$vs <- as.factor(c$vs)
```

## Simple regression

### Analysis

We fit a simple regression model using am as predictor and mpg as outcome.
We perform a t-test for validaing the hypothesis that mpg means differ across transmission type.
You will see in appendix that the residual error distrbution is almost normal, and the residual errors are equally spaned around 0.
 
```{r simpleregression-analysis, echo = FALSE}
lmCar_am <- lm(mpg ~ am, data = c)
coeflmCar_am <- summary(lmCar_am)$coefficients
coeflmCar_am
```


### Results
The p-value for the t-test is much less than 0.05 (cf appendix). The null hypothesis (equal means betFween automatic and manual transmission) is rejected.
**The transmission has an effect on mpg, manual transmission cars have globally a better mpg.**
According to this model, the gain between automatic and manual transmission is **`r round(coeflmCar_am[2, 1], 2)` miles / gallon**.
In order to adjust this gain value, we have now to consider a multivariate regression.

## Multivariate regression


### Select the accurate factors

We give below the standard error for each factor excluding am, ordered.
We will then perform a nested approach, including a factor after another, in this order.

```{r lmall-standarderrors}
lmAll = lm(mpg ~ . - am, data = c)
cv <- vif(lmAll)
stdErrors <- cv[order(cv[,3],decreasing=T),3]
dfcv <- data.frame()
dfcv <- rbind(dfcv, stdErrors)
colnames(dfcv) <- rownames(cv)
rownames(dfcv) <- c("stderror")
kable(dfcv)
```



```{r nested-models-1}
models <- c()
models <- vector("list", 9)
f1 <- lm(mpg ~ am, c); models[[1]] <- f1
f2 <- lm(mpg ~ am + disp, c); models[[2]] <- f2
f3 <- lm(mpg ~ am + disp + hp, c); models[[3]] <- f3
f4 <- lm(mpg ~ am + disp + hp + wt, c); models[[4]] <- f4
f5 <- lm(mpg ~ am + disp + hp + wt + cyl, c); models[[5]] <- f5
f6 <- lm(mpg ~ am + disp + hp + wt + cyl + qsec, c); models[[6]] <- f6
f7 <- lm(mpg ~ am + disp + hp + wt + cyl + qsec + vs, c); models[[7]] <- f7
f8 <- lm(mpg ~ am + disp + hp + wt + cyl + qsec + vs + gear, c); models[[8]] <- f8
f9 <- lm(mpg ~ am + disp + hp + wt + cyl + qsec + vs + gear + carb, c); models[[9]] <- f9
estimates <- c(); rse <- c()
for (mdl in models) {
    estimates <- c(estimates, round(mdl[1]$coefficients[2], 2))
    rse <- c(rse, round(sqrt(deviance(mdl)/df.residual(mdl)), 2))
}

dfModelClass <- data.frame()
dfModelClass <- rbind(dfModelClass, estimates); dfModelClass <- rbind(dfModelClass, rse)
colnames(dfModelClass) <- 1:9; rownames(dfModelClass) <- c("Estimate", "Resid. Std Error")
```

```{r nested-models-2}
a <- anova(f1, f2, f3, f4, f5, f6, f7, f8, f9)
dfa <- data.frame()
dfa <- rbind(dfa, round(a[[5]], 2))
dfa <- rbind(dfa, round(a[[6]], 4))
colnames(dfa) <- 1:9
rownames(dfa) <- c("F", "Pr(>F)")
```


```{r nested-models-3}
w <- waldtest(f1, f2, f3, f4, f5, f6, f7, f8, f9, test = "Chisq")
dfw <- data.frame()
dfw <- rbind(dfw, round(w[[3]], 2))
dfw <- rbind(dfw, round(w[[4]], 4))
colnames(dfw) <- 1:9
rownames(dfw) <- c("Chisq", "Pr(>Chisq)")
```

```{r model-select}
lmSel <- f5
# lmSelModel <- as.character(lmSel$call)
lmSelModel <- as.character(lmSel$call)
```


We select the model `r lmSelModel`. For this model, here is the summary.

```{r dataanalysis-multivariate, fig.height=8}
summary(lmSel)
```

### Results
With the model selected, the impact is `r lmSel$coefficients[2]`.

# Conclusion

The positive impact of manual transmission versus automatic transmission has been evaluated follwing 2 models:

1. Simple regression lm(mpg ~ am, c) : impact is 7.245
2. Multivariate regression lm(mpg ~ am + hp + cyl + wt, c) : impact is 1.8


# Appendix

```{r appendix-dependencies, ref.label='dependencies', eval=FALSE}

```

```{r appendix-loaddata, ref.label='loaddata', eval=FALSE}

```

```{r exploratoryanalysis, echo=FALSE, fig.height=3}
m <- mtcars
m$amfactor <- as.factor(m$am)
levels(m$amfactor) = c("auto", "manual")
g <- ggplot(m, aes(x = amfactor, y = mpg))
g <- g + geom_boxplot(aes(fill = amfactor))
g <- g + xlab("Transmission")
g <- g + ylab("Miles/(US) gallon")
g
```

```{r appendix-dataprocessing, ref.label='dataprocessing', eval=FALSE}

```

```{r appendix-exploratoryanalysis-2}
dim(mtcars)
kable(head(mtcars)) %>% kable_styling(bootstrap_options = c("striped"))
classes <- sapply(mtcars, class)
dfClasses <- data.frame()
dfClasses <- rbind(dfClasses, classes)
colnames(dfClasses) <- names(classes)
kable(dfClasses) %>% kable_styling(bootstrap_options = c("striped"))
table(mtcars$am)
?mtcars
```
```{r appendix-exploratoryanalysis, ref.label='exploratoryanalysis', eval=FALSE}

```


```{r appendix-simpleregression-analysis, fig.height=3}
t.test(c[c$am == "auto", c("mpg")], c[c$am == "manual", c("mpg")], alternative = "two.sided", var.equal = FALSE, paired = FALSE)
residCar_am <- residuals(lmCar_am)
fittedCar_am <- fitted.values(lmCar_am)
par(mfrow = c(1, 2))
plot(fittedCar_am, residCar_am, xlab = "Predicted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = "dashed")
plot(density(residCar_am), xlab = "Residuals", ylab = "Density", main = "")
par(mfrow = c(1, 1))
shapiro.test(lmCar_am$residuals)
```
### Covariances between factors
```{r appendix-factorcovariances, message=FALSE, fig.height=8, cache=TRUE}
ggpairs(c[, c("mpg", "am", "disp", "hp", "wt", "cyl", "qsec")],  
        lower = list(continuous = "smooth"),
        diag=list(continuous="bar"),
        # upper=list(), 
        axisLabels='show')
```

```{r appendix-lmall-standarderrors, ref.label="lmall-standarderrors", eval=FALSE}

```


```{r appendix-nested-models-1, ref.label="nested-models-1", eval=FALSE}

```

```{r appendix-nested-models-1-table}
kable(dfModelClass)
```

```{r appendix-nested-models-2, ref.label="nested-models-2", eval=FALSE}

```

```{r appendix-nested-models-2-table}
kable(dfa)
```

```{r appendix-nested-models-3, ref.label="nested-models-3", eval=FALSE}

```

```{r appendix-nested-models-3-table}
kable(dfw)
```

```{r appendix-multiregression-resid, fig.height=3}
resid <- residuals(lmSel)
fitted <- fitted.values(lmSel)
par(mfrow = c(1,2))
plot(density(resid), xlab = "Residuals", ylab = "Density", main = "Residual distribution")
plot(fitted, resid, xlab = "Predicted values", ylab = "Residuals")
abline(h = 0, col = "red", lty = "dashed")
shapiro.test(lmSel$residuals)
```

```{r appendix-multiregression-resid-2, fig.height=6}
par(mfrow = c(2,2))
plot(lmSel)
```

[http://github.com/vinceforce/blob/master/mpg_vs_transmission.Rmd]
