---
title: "Statistic Inference Course Project - Part 1"
author: "Vince Force"
date: "5 janvier 2018"
output:
  pdf_document: null
  md_document: null
  html_document: null
  word_document: default
mainfont: Arial
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loadlibaries, echo=FALSE}
library(ggplot2)
library(grid)
```


```{r multiplotfunction, echo=FALSE}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  
  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

```{r part1constants, echo=FALSE}
lambda <- 0.2
n <- 1000
nS <- 40
nSamples <- 1000
```

# Synopsis

We want to know if random exponential distribution follows the Central Limit Theorem (or not).

# Analysis
For illustrating data evidence, we draw two plots, using lambda = `r lambda` :

- histogram and density for a distribution of a large (n = `r n`) collection of random exponentials (no set.seed())
- histogram and density for the distribution of a large (nSamples = `r nSamples`) collection of averages of nS = `r nS` random exponentials (no set.seed()), and density of a normal distribution, with mean = 1/lambda (`r 1/lambda`) and sd = 1/lambda (`r 1/lambda`)

```{r Exercise 1, cache = TRUE, fig.height=3, message=FALSE, warning=FALSE}
expD <- rexp(n, lambda)
expDMean <- round(mean(expD), 2)
expDSd <- round(sd(expD), 2)
theme_update(plot.title=element_text(hjust=0.5, size=9),
             plot.subtitle=element_text(hjust=0.5, size=7))
df1 <- data.frame(expD)
g1 <- ggplot(df1, aes(x = expD)) + xlab("")
g1 <- g1 + geom_histogram(aes(y=..density..), color="black", fill="white")
g1 <- g1 + geom_density(alpha=.2, fill="#FF6666")
t1 <- paste("Histogram of an exponential distribution\n(", n, " elements)", sep = "")
g1 <- g1 + ggtitle(t1, subtitle = paste("mean : ", expDMean," | sd : ", expDSd, sep = ""))
expDSamples <- matrix(nrow = nSamples, ncol = nS)
for (i in 1:nSamples) expDSamples[i, ] <- rexp(nS, lambda)
expDSamplesMeans <- apply(expDSamples, MARGIN = 1, mean)
expDSamplesSds <- apply(expDSamples, MARGIN = 1, sd)
expDSamplesMeansMean <- round(mean(expDSamplesMeans), 2)
expDSamplesMeansSd <- round(sd(expDSamplesMeans), 2)
expDSamplesSdsMean <- round(mean(expDSamplesSds), 2)
df2 <- data.frame(expDSamplesMeans)
g2 <- ggplot(df2, aes(x = expDSamplesMeans)) + xlab("")
g2 <- g2 + geom_histogram(aes(y=..density..), color="black", fill="white")
g2 <- g2 + geom_density(alpha=.2, fill="#FF6666")
g2 <- g2 + stat_function(fun = dnorm, colour = "red",
            args = list(mean = 1/lambda, sd = (1/lambda) / sqrt(nS)))
st2 <- paste("mean of means : ", round(expDSamplesMeansMean, 2),
             " | mean of sds : ", round(expDSamplesSdsMean, 2),
             " | sd of means : ", round(expDSamplesMeansSd, 2),
             "\n1 / lambda = ", 1/lambda, " | (1 / lambda) / sqrt(", nS, ")) = ",
             round((1/lambda) / sqrt(nS), 2), sep = "")
g2 <- g2 + ggtitle(paste("Histogram of means distribution\n(means of ", nSamples,
                         " samples of ", nS, " elements each)", sep = ""), subtitle = st2)
multiplot(g1, g2, cols = 2)
```

# Results
On the figures above, we see on the figure on the left hand side that an exponential distrubution with `r n` elements does not follow a normal distribution, even if its mean and standard deviation are close to 1/lambda.

We see on the figure on the right on side (`r nSamples` iterations on a `r nS` elements samples) that :

- the density curve (in black) of the distribution of sample means is close to a normal density curve with mean = 1/lambda and sd = 1/lambda (in red)
- the mean of the distribution of sample means is very close to 1/lambda
- the mean of the distribution of sample sds is close to 1/lambda
- the standard deviation of the distribution of sample means is near from sqrt(nS) / lambda, where nS is the size of each sample

# Conclusion
We conclude that the random exponential distribution follows the Central Limit Theorem.

# Note on Rmd code
Some R code chunks have been masked for better presentation, including functions multiplot (for ggplot, found on googling)

Complete Rmd code can be found here

[https://github.com/vinceforce/datasciencecoursera/blob/master/StatisticInferenceCoursePoject-Part1.Rmd]