---
title: "Statistic Inference Course Project - Part 2"
author: "Vince Force"
date: "5 janvier 2018"
output:
  pdf_document: null
  md_document: null
  html_document: null
  word_document: default
mainfont: Arial
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loadlibaries, echo=FALSE}
library(ggplot2)
library(grid)
library(datasets)
```


```{r multiplotfunction, echo=FALSE}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  
  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

# Synopsis
We are looking for factors impacting teeth growth.

As we work with a quite small number of observations, we hope to be able to make a decision using inference. 

# Exploratory Analysis
After giving basic description (str, head, summary) of data, we plot barplots to have a look on data distribution following different groupings.

```{r part2constants, echo=FALSE}
data(ToothGrowth)
TG <- ToothGrowth
```
```{r part2datasummary}
str(TG)
head(TG)
summary(TG)
```


```{r Exercise 2, cache=TRUE, fig.height=3}
TG$dose <- as.factor(TG$dose)
g1 <- ggplot(TG, aes(x=supp, y=len)) + geom_boxplot()
g2 <- ggplot(TG, aes(x=dose, y=len)) + geom_boxplot()
multiplot(g1, g2, cols = 2)
```

On the left hand side, the two distributions differ, but not very much. The supp factor seems not to be very significant, even if means are close.

On the right hand side, the three distributions differ a lot one from the other. Dose factor seems much more significant.

Let us make some tests to investigate.
Considering the low number of observations (20), a t-distribution approach is convenient. We  assume equal variance in the various tests.

```{r poolsdfunction, echo=FALSE}
spf <- function(sx, sy, nx, ny) {
    mixprob <- (nx - 1) / (nx + ny - 2)
    sp <- sqrt((mixprob * sx^2 + (1 - mixprob) * sy^2))
}
```

# T-tests for supp

## Hypothesises
We assume H0 hypothesis is that means mOJ and mVC are equal, ie mOJ = mVC.
Ha is the hypothesis where moJ <> mVC.
First, we perform a power test, and the a t-test.

```{r Exercise 2 - t-tests-supp}
TGsuppOJ <- TG[TG$supp == "OJ", "len"]
TGsuppVC <- TG[TG$supp == "VC", "len"]
mOJ <- mean(TGsuppOJ)
mVC <- mean(TGsuppVC)
sOJ <- sd(TGsuppOJ)
sVC <- sd(TGsuppVC)
nOJ <- length(TGsuppOJ)
nVC <- length(TGsuppVC)
sp <- spf(sOJ, sVC, nOJ, nVC)
pow <- power.t.test(n = 30, delta = 0, sd = sp, alternative = "two.sided",
                    type = "two.sample")$power
paste("Power : ", pow)
t.test(TGsuppOJ, TGsuppVC, mu = 0, paired = FALSE, var.equal = TRUE,
       alternative = "two.sided")
```

## Results
As we can see, with a 0.05 significance level, the power is 2.5%, which is very low.
This is confirmed by the t-test, indicating :

- a p-value of 0.06, greater than the significance level
- a confidence interval containing the 0 value

We can not reject the H0 hypothesis.
That means that we can not be 95% sure that means differ.

## Conclusion
We can not tell, with this sample size, wether supp factor is significant on teeth growth or not.

# T-tests for dose

## Basic computations
```{r}
TGdose0.5 <- TG[TG$dose == 0.5, "len"]
TGdose1 <- TG[TG$dose == 1, "len"]
TGdose2 <- TG[TG$dose == 2, "len"]
m0.5 <- mean(TGdose0.5)
m1 <- mean(TGdose1)
m2 <- mean(TGdose2)
s0.5 <- sd(TGdose0.5)
s1 <- sd(TGdose1)
s2 <- sd(TGdose2)
n0.5 <- length(TGdose0.5)
n1 <- length(TGdose1)
n2 <- length(TGdose2)
```


### Hypothesises (dose = 0.5 versus dose = 1)
We assume H0 hypothesis is that means m0.5 and m1 are equal, ie m0.5 = m1.
Ha is the alternative hypothesis where m0.5 <> m1.
First, we perform a power test, and the a t-test.

```{r Exercise 2 - t-tests-dose-2}
sp <- spf(s0.5, s1, n0.5, n1)
pow <- power.t.test(n = 30, delta = m1 - m0.5, sd = sp, alternative = "two.sided",
             type = "two.sample")$power
paste("Power : ", pow)
t.test(TGdose1, TGdose0.5, mu = 0, paired = FALSE, var.equal = TRUE,
       alternative = "two.sided", conf.level = 0.95)
```

## Results (dose = 0.5 versus dose = 1)
As we can see, with a 0.05 significance level, the power is almost 100%, which is the best possible.

This is confirmed by the t-test, indicating :

- a p-value of 1.266e-07, much less than the significance level
- a 95% confidence interval [6.28 11.98] not containing the 0 value

We can reject the H0 hypothesis.

That means that we can be 95% sure that means differ. The dose factor seems to be much significant.

## Hypothesises (dose = 1 versus dose = 2)
We assume H0 hypothesis is that means m1 and m2 are equal, ie m1 = m2.
Ha is the hypothesis where m1 <> m2.
First, we perform a power test, and the a t-test.

```{r Exercise 2 - t-tests-dose-1}
sp <- spf(s1, s2, n1, n2)
pow <- power.t.test(n = 30, delta = m2 - m1, sd = sp, alternative = "two.sided",
             type = "two.sample")$power
paste("Power : ", pow)
t.test(TGdose2, TGdose1, mu = 0, paired = FALSE, var.equal = TRUE,
       alternative = "two.sided", conf.level = 0.95)
```

## Results (dose = 1 versus dose = 2)
As we can see, with a 0.05 significance level, the power is almost 100%, which is the best possible.
This is confirmed by the t-test, indicating :

- a p-value of 1.811e-05, much less than the significance level
- a 95% confidence interval [3.74 8.99] not containing the 0 value

We can reject the H0 hypothesis.
That means that we can be 95% sure that means differ. The dose factor seems to be much significant.

### Conclusion
A third test could have been done (comparing 0.5 with 2), but the result is obvious given the previous numeric comparison results and the boxplot.

We can tell, with this sample size, that the dose factor is significant : len mean is greater as the dose is greater. 

# Note on Rmd code
Some R code chunks have been masked for better presentation, including functions :

- multiplot (for ggplot, found on googling)
- spf (pool standard deviation with equal variance).

Complete Rmd code can be found here

[https://github.com/vinceforce/datasciencecoursera/blob/master/StatisticInferenceCoursePoject-Part2.Rmd]